Direct-style process creation on Linux

### 2020-05-11 research/paper/paper.tex
things

% \documentclass[sigplan]{acmart}
\documentclass{acmart}

@@

Direct-style is rare in general, but relatively common on academic operating systems.
% insert citations to direct-style papers here
Direct-style involves creating a new child process in a non-running state,
and then calling individual syscalls from the parent to initialize the child.
Direct-style does not constrain the calling process,
easily branches at any step of the initialization,
and can create any kind of process supported by the underlying system.

@@

This paper is organized in several sections.
In section \ref{background}, we provide background on fork-style, spawn-style, and direct-style process creation,
and consider their advantages and disadvantages in more depth.
In section \ref{examples} we show examples of usage of direct-style process creation in Python on Linux,
including use of process features which are difficult to use on Linux with \texttt{fork} or \texttt{posix\_spawn},
and briefly examine the user API.
In section \ref{implementation}, we explore our implementation in depth,
and discuss several difficult process-creation related details of \texttt{rsyscall}'s implementation.
In section \ref{evaluation}, we evaluate, somehow (TODO).
In section \ref{future_work},
we list several directions for future work related to direct-style process creation and to \texttt{rsyscall}.
In section \ref{conclusions}, we briefly and optimistically conclude.

@@

\section{Overview}\label{examples}
In most operating system APIs, including the ones available on Linux,
when a program makes a syscall, it implicitly operates on the current process.
To perform direct-style process creation on Linux,
we need a Linux API where we instead explicitly specify in which process we want to make a syscall.

@@

\begin{lstlisting}[float,language=Python,label={basic},caption={Creating a new process, changing CWD, and execing}]
child = local_thread.clone()
child.chdir("/dev")
child.execve("/bin/cat",
  ["cat", "./null"], local_thread.environ)
\end{lstlisting}
Listing \ref{basic} shows this in action in Python.
We create a new child thread using \texttt{clone} on \verb|local_thread|,
and receive back a capability to control it.
We pass no flags to use \texttt{clone}'s defaults
(which essentially match \texttt{fork}) to create the process.
We call \texttt{chdir} in the thread to change its working directory to \texttt{/dev}.
Then we call \texttt{execve},
passing the path of the \texttt{cat} executable,
the arguments to the new executable,
and the unchanged set of environment variables.
The name of the executable is passed as the first argument,
as is typical in Unix.
Calling \texttt{execve} consumes the capability,
and later use of this child thread capability will fail with an exception.

@@

The familiar and mundane nature of the API is important not just for the learning curve,
but also for expressivity.
Since the API is simply normal Linux system calls,
anything that could be expressed with fork and syscalls
can also be expressed with direct-style clone.
\section{Applications}
In this section,
we'll discuss several more advanced uses of processes,
using more sophisticated Linux features,
and accompany them with Python examples using direct-style process creation.

@@

By improving the process creation interface,
all these techniques can be used together, flexibly,
to more closely meet the specific needs of individual real-world applications.

\subsection{Abstraction through file descriptor passing}\label{fd_abstraction}
In Linux (and Unix in general),
a program can use file descriptor inheritance
to provide file descriptors created by the parent process to child processes.
Put simply,
a new Linux child process receives copies of all file descriptors in the parent process.
When calling \texttt{execve}, these file descriptors are normally closed automatically,
but selected file descriptors can be preserved over \texttt{execve}
by disabling the \texttt{CLOEXEC} flag on individual file descriptors.
We will discuss file descriptor inheritance in more detail in section \label{fdtables},
as part of our discussion of resource management in the implementation of \texttt{rsyscall}.

Most resources in Linux are managed through file descriptors,
so this allows the parent process to pass a wide variety of resources to the child process.
An executable program can be provided a file descriptor,
which can correspond to any file in any filesystem,
or to a network connection\cite{ucspi},
or a pipe connected to another process,
without the process being aware of what specific resource it is accessing.
In this way, we immediately know "for free"\cite{theoremsforfree}
that the behavior of the program does not depend on the name of the file,
or the hostname to it which the socket is connected,
or the executable running in the process on the other end of the pipe.
\begin{lstlisting}[float,language=Python,label={fds},caption={Passing down FDs}]
db_fd = local_thread.open("/var/db/database.db", O.RDWR)
child = local_thread.clone()
child_fd = child.inherit_fd(db_fd)
child_fd.fcntl(F.SETFD, 0)
child.execv("/bin/fooserver",
  ["fooserver", "--database-fd", str(int(child_fd))])
\end{lstlisting}
In Listing \ref{fds},
we first open a file with read-write permission.
Then we create a child thread,
which inherits all file descriptors from its parent.
We indicate that we want to operate on the child's copy of the file descriptor with \verb|child.inherit_fd|.
\verb|inherit_fd| only updates resource-tracking bookkeeping in our program;
it makes no system calls.
Then we disable \texttt{CLOEXEC} on the child's file descriptor,
which on Linux we can do by clearing the FD flags with \verb|fcntl(fd, F_SETFD, 0)|.
Finally, we execute some executable,
passing the file descriptor number as an argument;
when the executable starts up, that file descriptor number will already be pointing to this file descriptor.
We call the \texttt{execv} wrapper rather than \texttt{execve} as in listing \ref{basic}
so that we don't need to pass the (unmodified) environment to every exec call in these examples.
\subsection{Non-shared-memory concurrency}
Processes run concurrently,
which allows exploiting the parallelism of the underlying hardware.
Since processes don't share memory,
they can provide a less complex parallel programming environment
than shared-memory thread-based approaches.

A popular parallel programming environment is the Unix shell,
which obtains its parallelism by running multiple processes connected via pipes.
The Unix shell has a relatively constrained form of parallel processing,
but it's also possible to create more complex webs of parallel processes,
where, for example, one process might take multiple inputs over multiple pipes,
or produce multiple outputs.
\begin{lstlisting}[float,language=Python,label={pipe},caption={Creating a parallel processing pipeline}]
def as_argument(child, fd):
  child_fd = child.inherit_fd(fd)
  child_fd.fcntl(F.SETFD, 0) # unset CLOEXEC
  return str(int(child_fd)) # turn fd number to string

audio_pipe = local_thread.pipe()
video_pipe = local_thread.pipe()
source = local.thread.clone()
source.execve('/bin/source', ['source',
  '--audio-output', as_argument(source, audio_pipe.write),
  '--video-output', as_argument(source, video_pipe.write)])
video_sink = local.thread.clone()
video_sink.execve('/bin/video_sink', ['video_sink',
  '--video-input', as_argument(video_sink, video_pipe.read)])
audio_sink = local.thread.clone()
audio_sink.execve('/bin/audio_sink', ['audio_sink',
  '--audio-input', as_argument(audio_sink, audio_pipe.read)])
\end{lstlisting}
In listing \ref{pipe},
we execute a few programs in parallel,
connected by pipes.
We first create two pipes,
then inherit them down to several processes using a helper function, \texttt{as_argument},
which uses \verb|inherit_fd| and \texttt{fcntl} as described in section \ref{fd_abstraction}.
We exec each program passing some file descriptors as arguments.
The whole system then runs in parallel,
as normal child processes.
\subsection{Customization without explicit support}
While, ideally, all resources would be passed by file descriptor,[fn:capsicum]
Unix has a number of resources that are global,
such as the pid space or the mount table,
and which cannot be changed on a per-process basis.

@@

\begin{lstlisting}[float,language=Python,label={mount},caption={Overriding absolute path using a mount namespace}]
child = local.thread.clone(CLONE.NEWNS)
child.mount("/home/foo/custom_foo.conf", "/etc/foo.conf", "", MS.BIND, "")
child.execv('/bin/fooserver', ['fooserver'])
\end{lstlisting}

One classic form of customization enabled by namespaces
is overriding files at paths that were hardcoded into the program.
In listing \ref{mount},
we create a new child thread as normal,
but for the first time,
pass an argument to \texttt{clone}.
\verb|CLONE_NEWNS| causes clone to create the process in a new mount namespace,
which allows us to use the \texttt{mount} system call
to manipulate the filesystem tree in a way that only this process will see.
We call mount, passing \texttt{MS.BIND}, to make the file \verb|/home/foo/custom_foo.conf|
appear at the path \verb|/etc/foo.conf|;
this is known as a bind mount.
Then we execute some executable,
which will see the \verb|/etc/foo.conf| as we specified.
\subsection{Sandboxing}
At process creation time,
we can not only pass resources and customize the process's environment,
we can also deny the process access to resources that it otherwise gets by default.
This is a key part of creating a secure sandbox for potentially malicious code.

@@

\begin{lstlisting}[float,language=Python,label={unmount},caption={Unmount all and run executable via fexec}]
thread = local_thread.clone(CLONE.NEWNS)
executable_fd = child.open("/bin/static_executable", O.RDONLY)
db_fd = child.open("/var/db/database.db", O.RDWR)
child.umount("/", MNT.DETACH)
db_fd.fcntl(F.SETFD, 0)
child.fexec(executable_fd, ["static_executable", "--database-fd", str(int(db_fd))])
\end{lstlisting}

In listing \ref{unmount}
we create a new child thread,
again in a new mount namespace using \texttt{CLONE.NEWNS}.
We open the executable that we'll later exec in this child thread,
as well as another file as in listing \ref{fds},
since we won't be able to do exec or open files normally after the next step.
Then we use \texttt{umount},
passing \texttt{MNT>DETACH} to perform a recursive unmount of the entire filesystem tree,
removing it all from the view of this process.
As in listing \ref{fds}, we unset the \texttt{CLOEXEC} flag from the database file descriptor,
so that it can be inherited across \texttt{exec};
we don't need to call \verb|inherit_fd| since, in this case, the database fd was opened directly from the child.
We then exec the file we opened earlier, using \texttt{fexec},
which allows execing a file descriptor,
and also pass the database file descriptor as an argument,
as in listing \ref{fds}.
(The underlying system call for \texttt{fexec} is \texttt{execveat};
this \texttt{fexec} is a helper function.)
Note that this executable must be statically linked,
or it wouldn't work in an empty filesystem namespace,
with no libraries to dynamically link against.

@@

Even if the process needs additional resources,
those can be explicitly passed down through file descriptor passing,
as we do here with the database file descriptor.
This is the essence of capability-based security,
which Linux can increasingly closely approximate.

\subsection{Nested clone and pid namespaces}
Process-local resource can also be used to control the lifetime of resources used by that process.
Many Linux resources are not automatically cleaned up on process exit;
a poorly coded program may allocate global resources
and not ensure that they are cleaned up,
leaving behind unuseds litter on the system.

@@

\begin{lstlisting}[float,language=Python,label={pidns},caption={Nested clone and pid namespace}]
init = local_thread.clone(CLONE.NEWPID)
grandchild = init.clone()
grandchild.execv('/bin/fooserver', ['fooserver'])
\end{lstlisting}

In listing \ref{pidns},
we create a child process which we call \texttt{init},
passing \verb|CLONE_NEWPID| to create it in a new pid namespace.
To create another process in the pid namespace,
we clone again, this time from \texttt{init};
note that this is the first time we've cloned from one of our child processes.
We exec in the grandchild,
and we can monitor the grandchild process from \texttt{init},
just as we would monitor \texttt{init} from its parent process.

@@

In either case, when \texttt{init} dies,
the pid namespace will be destroyed,
and \texttt{grandchild} will be cleaned up.

\subsection{Shared file descriptor tables}
A library can create a child process to perform a useful service.
This can sometimes be done more easily when the child process shares file descriptors with the parent.
\begin{lstlisting}[float,language=Python,label={fuse},caption={Shared file descriptor tables}]
ns_child = local_thread.clone(CLONE.FILES|CLONE.NEWNS)
server_child = ns_child.clone()
await server_child.execve('/bin/foofs', ['foofs', "--mount-at", "/"])
fd = ns_child.open("/foo/bar", O.RDONLY)
parent_fd = local_thread.use_fd(fd)
\end{lstlisting}
In listing \ref{fuse},
we first create a child process, \verb|ns_child|, in a new mount namespace with \verb|CLONE_NEWNS|,
and for the first time, also pass \verb|CLONE_FILES|,
which causes the file descriptor table to be shared between the parent process and the child process.
We create a grandchild off of \verb|ns_child|,
and use it to exec a FUSE filesystem,
which will appear only in the mount namespace of \verb|ns_child| and its descendents.
Then we can open FUSE files in \verb|ns_child|
and use those files in the parent process,
through the shared file descriptor table.
\section{Implementation}\label{implementation}
\subsection{Basics about rsyscall}
Our main need for implementing direct-style process creation
is a robust system for cross-process syscalls.
The rsyscall project provides this.

@@

An additional argument to the \texttt{execve} syscall
which allows specifying an explicit list of file descriptor numbers to inherit despite the \texttt{CLOEXEC} flag being set
would allow programs to avoid this pointless behavior of un-setting and then immediately re-setting \texttt{CLOEXEC}.
It would also, more significantly, allow programs to inherit file descriptors across \texttt{execve}
while the file descriptor table is shared;
unsetting \texttt{CLOEXEC} while the file descriptor table is shared will cause race conditions,
because the other processes sharing the file descriptor table might call \texttt{execve} at any time,
and so the file descriptor table must first be unshared before unsetting \texttt{CLOEXEC} and calling \texttt{execve}.
\section{Evaluation}\label{evaluation}
\subsection{performance ideas}
Can mmap a bunch of stuff in a process and then create some children,
and show that without having to copy page tables (due to \verb|CLONE_VM|), it's faster.

Can make the classic chart of fork vs glibc's \verb|posix_spawn| vs direct-style.
\subsection{abstraction ideas}
Implementing Python's popen/subprocess with rsyscall?
But that's not fair, theirs is portable, we'll be way simpler for free.

Rewrite some existing system with rsyscall?
Write a shell?
\section{Future work}\label{future_work}
\subsection{Other applications of rsyscall}
Most avenues of future work focus on rsyscall.
rsyscall was not developed solely for the purpose of this paper,
and it has many uses unrelated to direct-style process creation,
such as asynchronous system calls, exceptionless system calls\cite{flexsc}, cross-host operations, among others.
We are actively exploring such applications,
as well as broadening rsyscall's language support.
\subsection{Kernel support}
rsyscall's cross-process syscalls can be performed entirely in userspace,
which has substantial benefits for deployability.
Nevertheless, direct kernel support for creating a stub process and performing syscalls in the context of that process
may provide efficiency benefits, as well as reducing userspace-visible complexity.

@@

\bibliographystyle{ACM-Reference-Format}
\bibliography{bibliography}

\end{document}

>>>>

\documentclass[sigplan]{acmart}
% \documentclass{acmart}

@@

Direct-style is rare in general, but relatively common on academic operating systems.
\cite{keykos}\cite{sel4}\cite{exokernel}\cite{fuschia}\cite{singularity}
Direct-style involves creating a new child process in a non-running state,
and then calling individual syscalls from the parent to initialize the child.
Direct-style does not constrain the calling process,
easily branches at any step of the initialization,
and can create any kind of process supported by the underlying system.

@@

This paper is organized in several sections.
In section \ref{background}, we provide background on fork-style, spawn-style, and direct-style process creation,
and consider their advantages and disadvantages in more depth.
In section \ref{overview} we discuss the design of the API,
including our usage of \texttt{rsyscall}.
In section \ref{applications} we show examples of usage of direct-style process creation in Python on Linux,
including use of process features which are difficult to use on Linux with \texttt{fork} or \texttt{posix\_spawn}.
In section \ref{implementation}, we explore our implementation in depth,
and discuss several difficult process-creation related details of \texttt{rsyscall}'s implementation.
In section \ref{evaluation}, we evaluate, somehow (TODO).
In section \ref{future_work},
we list several directions for future work related to direct-style process creation and to \texttt{rsyscall}.
In section \ref{conclusions}, we briefly and optimistically conclude.

@@

\section{Overview}\label{overview}
In most operating system APIs, including the ones available on Linux,
when a program makes a syscall, it implicitly operates on the current process.
To perform direct-style process creation on Linux,
we need a Linux API where we instead explicitly specify in which process we want to make a syscall.

@@

\begin{lstlisting}[float=*,language=Python,label={basic},caption={Creating a new process, changing CWD, and execing}]
child = local_thread.clone()
child.chdir("/dev")
child.execve("/bin/cat",
  ["cat", "./null"], local_thread.environ)
\end{lstlisting}
Listing \ref{basic} shows this in action in Python.
We create a new child thread using \texttt{clone} on \verb|local_thread|,
and receive back a capability to control it.
We pass no flags to use \texttt{clone}'s defaults
(which essentially match \texttt{fork}) to create the process.
We call \texttt{chdir} in the thread to change its working directory to \texttt{/dev}.
Then we call \texttt{execve},
passing the path of the \texttt{cat} executable,
the arguments to the new executable,
and the unchanged set of environment variables.
The name of the executable is passed as the first argument,
as is typical in Unix.
Calling \texttt{execve} consumes the capability,
and later use of this child thread capability will fail with an exception.

@@

The familiar and mundane nature of the API is important not just for the learning curve,
but also for expressivity.
Since the API is simply normal Linux system calls,
anything that could be expressed with fork and syscalls
can also be expressed with direct-style clone.
\section{Applications}\label{applications}
In this section,
we'll discuss several more advanced uses of processes,
using more sophisticated Linux features,
and accompany them with Python examples using direct-style process creation.

@@

By improving the process creation interface,
all these techniques can be used together, flexibly,
to more closely meet the specific needs of individual real-world applications.
\subsection{Abstraction through file descriptor passing}\label{fd_abstraction}
In Linux (and Unix in general),
a program can use file descriptor inheritance
to provide file descriptors created by the parent process to child processes.
Put simply,
a new Linux child process receives copies of all file descriptors in the parent process.
When calling \texttt{execve}, these file descriptors are normally closed automatically,
but selected file descriptors can be preserved over \texttt{execve}
by disabling the \texttt{CLOEXEC} flag on individual file descriptors.
We will discuss file descriptor inheritance in more detail in section \label{fdtables},
as part of our discussion of resource management in the implementation of \texttt{rsyscall}.

Most resources in Linux are managed through file descriptors,
so this allows the parent process to pass a wide variety of resources to the child process.
An executable program can be provided a file descriptor,
which can correspond to any file in any filesystem,
or to a network connection\cite{ucspi},
or a pipe connected to another process,
without the process being aware of what specific resource it is accessing.
In this way, we immediately know "for free"\cite{theoremsforfree}
that the behavior of the program does not depend on the name of the file,
or the hostname to it which the socket is connected,
or the executable running in the process on the other end of the pipe.

\begin{lstlisting}[float=*,language=Python,label={fds},caption={Passing down FDs}]
db_fd = local_thread.open("/var/db/database.db", O.RDWR)
child = local_thread.clone()
child_fd = child.inherit_fd(db_fd)
child_fd.fcntl(F.SETFD, 0)
child.execv("/bin/fooserver",
  ["fooserver", "--database-fd", str(int(child_fd))])
\end{lstlisting}
In Listing \ref{fds},
we first open a file with read-write permission.
Then we create a child thread,
which inherits all file descriptors from its parent.
We indicate that we want to operate on the child's copy of the file descriptor with \verb|child.inherit_fd|.
\verb|inherit_fd| only updates resource-tracking bookkeeping in our program;
it makes no system calls.
Then we disable \texttt{CLOEXEC} on the child's file descriptor,
which on Linux we can do by clearing the FD flags with \verb|fcntl(fd, F_SETFD, 0)|.
Finally, we execute some executable,
passing the file descriptor number as an argument;
when the executable starts up, that file descriptor number will already be pointing to this file descriptor.
We call the \texttt{execv} wrapper rather than \texttt{execve} as in listing \ref{basic}
so that we don't need to pass the (unmodified) environment to every exec call in these examples.
\subsection{Non-shared-memory concurrency}
Processes run concurrently,
which allows exploiting the parallelism of the underlying hardware.
Since processes don't share memory,
they can provide a less complex parallel programming environment
than shared-memory thread-based approaches.

A popular parallel programming environment is the Unix shell,
which obtains its parallelism by running multiple processes connected via pipes.
The Unix shell has a relatively constrained form of parallel processing,
but it's also possible to create more complex webs of parallel processes,
where, for example, one process might take multiple inputs over multiple pipes,
or produce multiple outputs.

\begin{lstlisting}[float=*,language=Python,label={pipe},caption={Creating a parallel processing pipeline}]
def as_argument(child, fd):
  child_fd = child.inherit_fd(fd)
  child_fd.fcntl(F.SETFD, 0) # unset CLOEXEC
  return str(int(child_fd)) # turn fd number to string

audio_pipe = local_thread.pipe()
video_pipe = local_thread.pipe()
source = local.thread.clone()
source.execve('/bin/source', ['source',
  '--audio-output', as_argument(source, audio_pipe.write),
  '--video-output', as_argument(source, video_pipe.write)])
video_sink = local.thread.clone()
video_sink.execve('/bin/video_sink', ['video_sink',
  '--video-input', as_argument(video_sink, video_pipe.read)])
audio_sink = local.thread.clone()
audio_sink.execve('/bin/audio_sink', ['audio_sink',
  '--audio-input', as_argument(audio_sink, audio_pipe.read)])
\end{lstlisting}
In listing \ref{pipe},
we execute a few programs in parallel,
connected by pipes.
We first create two pipes,
then inherit them down to several processes using a helper function, \verb|as_argument|,
which uses \verb|inherit_fd| and \texttt{fcntl} as described in section \ref{fd_abstraction}.
We exec each program passing some file descriptors as arguments.
The whole system then runs in parallel,
as normal child processes.
\subsection{Customization without explicit support}
While, ideally, all resources would be passed by file descriptor,[fn:capsicum]
Unix has a number of resources that are global,
such as the pid space or the mount table,
and which cannot be changed on a per-process basis.

@@

\begin{lstlisting}[float=*,language=Python,label={mount},caption={Overriding absolute path using a mount namespace}]
child = local.thread.clone(CLONE.NEWNS)
child.mount("/home/foo/custom_foo.conf", "/etc/foo.conf", "", MS.BIND, "")
child.execv('/bin/fooserver', ['fooserver'])
\end{lstlisting}
One classic form of customization enabled by namespaces
is overriding files at paths that were hardcoded into the program.
In listing \ref{mount},
we create a new child thread as normal,
but for the first time,
pass an argument to \texttt{clone}.
\verb|CLONE_NEWNS| causes clone to create the process in a new mount namespace,
which allows us to use the \texttt{mount} system call
to manipulate the filesystem tree in a way that only this process will see.
We call mount, passing \texttt{MS.BIND}, to make the file \verb|/home/foo/custom_foo.conf|
appear at the path \verb|/etc/foo.conf|;
this is known as a bind mount.
Then we execute some executable,
which will see the \verb|/etc/foo.conf| as we specified.
\subsection{Sandboxing}
At process creation time,
we can not only pass resources and customize the process's environment,
we can also deny the process access to resources that it otherwise gets by default.
This is a key part of creating a secure sandbox for potentially malicious code.

@@

\begin{lstlisting}[float=*,language=Python,label={unmount},caption={Unmount all and run executable via fexec}]
thread = local_thread.clone(CLONE.NEWNS)
executable_fd = child.open("/bin/static_executable", O.RDONLY)
db_fd = child.open("/var/db/database.db", O.RDWR)
child.umount("/", MNT.DETACH)
db_fd.fcntl(F.SETFD, 0)
child.fexec(executable_fd, ["static_executable", "--database-fd", str(int(db_fd))])
\end{lstlisting}
In listing \ref{unmount}
we create a new child thread,
again in a new mount namespace using \texttt{CLONE.NEWNS}.
We open the executable that we'll later exec in this child thread,
as well as another file as in listing \ref{fds},
since we won't be able to do exec or open files normally after the next step.
Then we use \texttt{umount},
passing \texttt{MNT.DETACH} to perform a recursive unmount of the entire filesystem tree,
removing it all from the view of this process.
As in listing \ref{fds}, we unset the \texttt{CLOEXEC} flag from the database file descriptor,
so that it can be inherited across \texttt{exec};
we don't need to call \verb|inherit_fd| since, in this case, the database fd was opened directly from the child.

We then exec the file we opened earlier, using \texttt{fexec},
which allows execing a file descriptor,
and also pass the database file descriptor as an argument,
as in listing \ref{fds}.
(The underlying system call for \texttt{fexec} is \texttt{execveat};
this \texttt{fexec} is a helper function.)
Note that this executable must be statically linked,
or it wouldn't work in an empty filesystem namespace,
with no libraries to dynamically link against.

@@

Even if the process needs additional resources,
those can be explicitly passed down through file descriptor passing,
as we do here with the database file descriptor.
This is the essence of capability-based security,
which Linux can increasingly closely approximate.
\subsection{Nested clone and pid namespaces}
Process-local resource can also be used to control the lifetime of resources used by that process.
Many Linux resources are not automatically cleaned up on process exit;
a poorly coded program may allocate global resources
and not ensure that they are cleaned up,
leaving behind unuseds litter on the system.

@@

\begin{lstlisting}[float=*,language=Python,label={pidns},caption={Nested clone and pid namespace}]
init = local_thread.clone(CLONE.NEWPID)
grandchild = init.clone()
grandchild.execv('/bin/fooserver', ['fooserver'])
\end{lstlisting}
In listing \ref{pidns},
we create a child process which we call \texttt{init},
passing \verb|CLONE_NEWPID| to create it in a new pid namespace.
To create another process in the pid namespace,
we clone again, this time from \texttt{init};
note that this is the first time we've cloned from one of our child processes.
We exec in the grandchild,
and we can monitor the grandchild process from \texttt{init},
just as we would monitor \texttt{init} from its parent process.

@@

In either case, when \texttt{init} dies,
the pid namespace will be destroyed,
and \texttt{grandchild} will be cleaned up.
\subsection{Shared file descriptor tables}
A library can create a child process to perform a useful service.
This can sometimes be done more easily when the child process shares file descriptors with the parent.

\begin{lstlisting}[float=*,language=Python,label={fuse},caption={Shared file descriptor tables}]
ns_child = local_thread.clone(CLONE.FILES|CLONE.NEWNS)
server_child = ns_child.clone()
await server_child.execve('/bin/foofs', ['foofs', "--mount-at", "/"])
fd = ns_child.open("/foo/bar", O.RDONLY)
parent_fd = local_thread.use_fd(fd)
\end{lstlisting}
In listing \ref{fuse},
we first create a child process, \verb|ns_child|, in a new mount namespace with \verb|CLONE_NEWNS|,
and for the first time, also pass \verb|CLONE_FILES|,
which causes the file descriptor table to be shared between the parent process and the child process.
We create a grandchild off of \verb|ns_child|,
and use it to exec a FUSE filesystem,
which will appear only in the mount namespace of \verb|ns_child| and its descendents.
Then we can open FUSE files in \verb|ns_child|
and use those files in the parent process,
through the shared file descriptor table.
\section{Implementation}\label{implementation}
\subsection{Basics about rsyscall}
Our main need for implementing direct-style process creation
is a robust system for cross-process syscalls.
The rsyscall project provides this.

@@

An additional argument to the \texttt{execve} syscall
which allows specifying an explicit list of file descriptor numbers to inherit despite the \texttt{CLOEXEC} flag being set
would allow programs to avoid this pointless behavior of un-setting and then immediately re-setting \texttt{CLOEXEC}.
It would also, more significantly, allow programs to inherit file descriptors across \texttt{execve}
while the file descriptor table is shared;
unsetting \texttt{CLOEXEC} while the file descriptor table is shared will cause race conditions,
because the other processes sharing the file descriptor table might call \texttt{execve} at any time,
and so the file descriptor table must first be unshared before unsetting \texttt{CLOEXEC} and calling \texttt{execve}.
\section{Evaluation}\label{evaluation}
    We quantitatively answer the following questions:

    - Does direct-style process creation on Linux allow for simpler programs than the alternatives?
      We implement and compare similar functionality using both direct-style and fork-style.\ref{ease}

    - Does a clone-based implementation avoid the well-known performance overheads of fork?
      We evaluate clone against fork on several microbenchmarks.\ref{microbench}

    - Does a high-level Python direct-style process creation interface
      have acceptable performance relative to the Python standard library's spawn-style interface?
      We benchmark direct-style process creation against spawn-style.\ref{subprocess_run}
\subsection{Ease of programming with direct-style}\label{ease}
Implementing Python's popen/subprocess with rsyscall?
But that's not fair, theirs is portable, we'll be way simpler for free.

Rewrite some existing system with rsyscall?
Write a shell?
\subsection{Clone versus fork}\label{microbench}
Can mmap a bunch of stuff in a process and then create some children,
and show that without having to copy page tables (due to \verb|CLONE_VM|), it's faster.

Can make the classic chart of fork vs glibc's \verb|posix_spawn| vs direct-style.
\subsection{High-level direct-style versus spawn-style}\label{subprocess_run}
Compare clone/exec with rsyscall against Python's subprocess.run.
\section{Related work}\label{related_work}
% TODO all this is just slapped onto the page without much thought, needs revising
\subsection{Direct-style process creation}
As we discuss in section \ref{background},
many academic operating systems have direct-style process creation.
We are not aware of any previous instances of direct-style process creation in a Unix-like environment.

The closest related work known to us
is efforts to build Unix compatibility environments, including fork-style process creation,
on academic operating systems which natively use direct-style process creation.\cite{exokernel}
Such compatibility environments could theoretically be bypassed to perform direct-style process creation
using the underlying primitives,
but regular Unix system calls could not be used to create new Unix processes in a direct-style way
in such an environment.
\subsection{Remote system calls}
There are a large number of Unix-based systems which do something called "remote system calls".
Most such systems do not allow a single program to manipulate multiple processes.
Those systems that do allow manipulating multiple processes
are generally oriented towards debugging and introspection,
and are unsuitable for a general purpose system.

Many systems use system call forwarding to implement migration in a computing cluster.
HTCondor\cite{condor} and Popcorn\cite{popcorn} are two examples.
In both systems, processes can be live-migrated between hosts;
when this occurs, the system will automatically forward IO-related system calls
back to the original host.

Some systems intercept system calls and forward them to another process to be implemented
as part of a virtualization system.
gvisor\cite{gvisor} and ViewOS\cite{viewos} are two examples.
In both systems,
system calls made in one process are intercepted,
and serviced by another process.
Also relevant is seccomp system call interception,
which can be used to intercept system calls and emulate them in another process.

Some systems forward system calls elsewhere for asynchronous execution.
FlexSC\cite{flexsc} is one example.
System calls were forwarded to dedicated system call execution threads.
The recently added Linux feature \verb|io_uring| is, in some sense, another example;
it supports sending a subset of system calls to the kernel,
which executes them asynchronously,
rather than blocking the thread.

Some systems call system calls in other processes for debugging or introspection purposes.
strace, gdb, and CRIU\cite{criu} are examples of this.
These systems typically uses ptrace,
which is capable of operating on multiple processes at once.
Unfortunately, as discussed in section \ref{implementation},
ptrace is unsuitable as a mechanism for a general purpose system since it is not re-entrant;
a process which is already being ptraced cannot be ptraced again,
so using ptrace as a cross-process system call mechanism would disallow use of strace or debuggers.
\subsection{Process capabilities}
We mentioned the concept of process capabilities while discussing the design of rsyscall in section \ref{overview}.
The Capsicum system provides something called process descriptors\cite{capsicum},
which is a file descriptor handle for a process.
A similar feature has been recently added to Linux in the form of the pidfd API\cite{pidfd}.
These notions of process capability are quite limited, however;
pidfds and process descriptors only allow sending signals to a process or waiting for its death,
rather than exercising full control over the process.
\subsection{Capability-secure libc replacements}
The Capsicum system provides a set of system calls
which can be used to provide a capability-secure sandbox.\cite{capsicum}
Latter efforts\cite{oblivious}, including CloudABI\cite{cloudabi} and WASI\cite{wasi},
are focused primarily on backwards compatibility with the existing POSIX environment, and ease of porting existing code.
Programs ported to the capability-secure environment
are typically launched with an accompanying spawn-style wrapper.
This means that capability-secure processes are created primarily by an API
that is distinct from the capability-secure syscall API they have access to,
which makes it harder for a capability-secure process to spawn further capability-secure processes of its own.
Direct-style process creation could make it easier to create capability-secure processes 
from the capability-secure syscall API,
and reduce the need for a separate spawn-style wrapper.

Also related is the PLASH, ``Principle of Least Authority Shell''\cite{plash}.
PLASH provides a spawn-style API (in the form of a shell)
to launch processes in a capability-secure environment.
PLASH, like all spawn-style APIs, abstracts over the native Linux environment,
and is therefore limited in what kind of processes it can create.
\section{Future work}\label{future_work}
\subsection{Other applications of rsyscall}
Most avenues of future work focus on rsyscall.
rsyscall was not developed solely for the purpose of this paper,
and it has many uses unrelated to direct-style process creation,
such as asynchronous system calls, exceptionless system calls\cite{flexsc}, cross-host operations, among others.
We are actively exploring such applications,
as well as broadening rsyscall's language support.
\subsection{Kernel support}
rsyscall's cross-process syscalls can be performed entirely in userspace,
which has substantial benefits for deployability.
Nevertheless, direct kernel support for creating a stub process and performing syscalls in the context of that process
may provide efficiency benefits, as well as reducing userspace-visible complexity.

@@

\bibliographystyle{ACM-Reference-Format}
\bibliography{bibliography}
\end{document}
